---
title: "04. Plotting multiple moving averages"
format: html
editor: visual
---

## Portraying exponenital moving averages in time series data

### Set things up and pull in our carbon dioxide data

We need to add a couple of packages for helping to calculate moving averages.

```{r}
rm(list = ls())
library(tidyverse)
library(tidyquant) #for moving averages
library(gridExtra) #to arrange plots
library(here)
```

Pull in our carbon dioxide data set

```{r}
co2 <-read.csv(here("Data/time-series-data/mauna-loa-carbon-dioxide.csv"))
```

Let's make the date into a date format for plotting

```{r}
co2$date <- ymd(co2$date)
```

## Plotting multiple moving averages on the same plot.

It turns out that the `tidyquant` package can calculate more than one type of moving average, and, because it has a moving average `geom` for use in ggplot, we can add different moving averages a layers to the same plot. 

Let's take a look.

We use ggplot to make the basic plot and then add some special geoms from `tidyquant`.

```{r}
co2_plot <- co2 %>% ggplot(aes(date, carbon.dioxide.ppm))+
  geom_line(color = "blue", size = 0.75)+ #original data
  geom_ma(ma_fun = SMA, #specifies simple moving average,
          n = 365, #specifies number of days
          color = "red", size = 1)+
  geom_ma(ma_fun = EMA, #specify type of moving average
          n = 1095, #specify the number of most recent points for the moving average
          color = "green", size = 1)+
  labs(title= "Plotting multiple moving averages", x = "Date", y = "Atmospheric CO2, ppm")+
  theme_bw()
co2_plot
```

We can mabe see the differences better if we zoom in:

```{r}
co2 %>% filter(Year > 2013) %>% 
  ggplot(aes(date, carbon.dioxide.ppm))+
  geom_line(color = "blue", size = 0.75)+ #original data
  geom_ma(ma_fun = SMA, #specifies simple moving average,
          n = 365, #specifies number of days
          color = "red", size = 1)+
  geom_ma(ma_fun = EMA, #specify type of moving average
          n = 1095, #specify the number of most recent points for the moving average
          color = "green", size = 1)+
  labs(title= "Plotting multiple moving averages", x = "Date", y = "Atmospheric CO2, ppm")+
  theme_bw()
```

Notice that the exponential moving average trend line appears somewhat "beneath" the center of the point cloud (where the simple moving average red line is located) - this is due to the difference in how a EMA is calculated.

